---
title: "What's New in tidyvpc?"
output: rmarkdown::html_vignette
resource_files:
  - img/vpc_results_overview.gif
vignette: >
  %\VignetteIndexEntry{What's New in tidyvpc?}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, warning = FALSE, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 9, fig.height = 6, out.width = 640, warning = FALSE)
```

## tidyvpc 1.5.0

### Plot Percentage of BLQ and/or ALQ

For VPC with censoring, users can supply additional arguments to `plot.tidyvpcobj` e.g., `censoring.type` (options are `'none'`, `'blq'`, `'alq'`, or `'both'`, defaults to `'none'`) and `censoring.output` (options are `'grid'` or `'list'`, defaults to `'grid'`).

If `censoring.output = 'grid'`, the plots will be arranged into single grid plot. Users may pass additional arguments via ellipsis to `egg::ggarrange` e.g., `nrow = 1`, `ncol = 2` in order to customize plots in grid arrangement.

If `censoring.output = 'list'`, the resulting plots will be returned individually as elements in list.

Example usage is below:

```{r}
library(tidyvpc)
obs_data <- obs_data[MDV == 0]
sim_data <- sim_data[MDV == 0]
obs_data$LLOQ <- obs_data[, ifelse(STUDY == "Study A", 50, 25)]
obs_data$ULOQ <- obs_data[, ifelse(STUDY == "Study A", 125, 100)]

vpc <- observed(obs_data, x = TIME, y = DV) |>
  simulated(sim_data, y = DV) |>
  censoring(blq = DV < LLOQ, lloq = LLOQ,  alq = DV > ULOQ, uloq = ULOQ) |>
  stratify(~ STUDY) |>
  binning(bin = NTIME) |>
  vpcstats(qpred = c(0.1, 0.5, 0.9))
```

If blq data, users may supply `censoring.type = "blq"`:

```{r}
plot(vpc, censoring.type = "blq", censoring.output = "grid", facet.scales = "fixed")
```

If alq data, users may supply `censoring.type = "alq"`:

```{r}
plot(vpc, censoring.type = "alq", censoring.output = "grid", ncol = 2, nrow = 1)
```

If both blq and alq data, users may supply `censoring.type = "both"`

```{r}
vpc_plots <- plot(vpc, censoring.type = "both", censoring.output = "list")
```

By default, when `censoring.tidyvpcobj` is used, no percentage blq/alq plots will be returned e.g., default for `censoring.type = 'none'`. If users specify `censoring.type='both'` and only blq censoring was performed, for example, they will receive an error stating e.g., `pctalq data.frame was not found in tidyvpcobj. Use censoring() to create censored data for plotting alq`.

## tidyvpc 1.4.0

### Additional Binning Methods

The following additional binning methods from `classInt` have been made available in `tidyvpc`. 

see `?classIntervals` 'style' descriptions for applicable arguments for each selected binning method.

```{r}
library(tidyvpc)
obs_data <- obs_data[MDV == 0]
sim_data <- sim_data[MDV == 0]
```


#### `headtails`

```{r}
observed(obs_data, x = TIME, y = DV) |>
  simulated(sim_data, y = DV) |>
  binning(bin = "headtails") |>
  plot()
```

Including additional `thr` argument.

```{r}
observed(obs_data, x = TIME, y = DV) |>
  simulated(sim_data, y = DV) |>
  binning(bin = "headtails", thr = 0.55) |>
  plot()
```


#### `maximum`

```{r}
observed(obs_data, x = TIME, y = DV) |>
  simulated(sim_data, y = DV) |>
  binning(bin = "maximum") |>
  plot()

```


Including additional `nbins` argument.

```{r}
observed(obs_data, x = TIME, y = DV) |>
  simulated(sim_data, y = DV) |>
  binning(bin = "maximum", nbins = 7) |>
  plot()
```


#### `box`

```{r}
observed(obs_data, x = TIME, y = DV) |>
  simulated(sim_data, y = DV) |>
  binning(bin = "box") |>
  plot()
```

Including additional `iqr_mult` and `type` argument.

```{r}
observed(obs_data, x = TIME, y = DV) |>
  simulated(sim_data, y = DV) |>
  binning(bin = "box", iqr_mult = 4) |>
  plot()

# additional (quantile) type arg
observed(obs_data, x = TIME, y = DV) |>
  simulated(sim_data, y = DV) |>
  binning(bin = "box", type = 3) |>
  plot()

```

### Additional flexibility for binless() + predcorrect()

Users may now execute `predcorrect()` either before, or after calling `binless(loess.ypc=TRUE)`. Previously, you were required to execute `predcorrect()` before `binless(loess.ypc=TRUE)`, otherwise you'd receive an error.

The following code below produces equivalent output:

```{r eval=FALSE}
observed(obs_data, x = TIME, y = DV ) |>
  simulated(sim_data, y = DV) |>
  stratify(~ GENDER) |>
  predcorrect(pred=PRED) |> #before binless()
  binless(loess.ypc=TRUE) |>
  vpcstats() |>
  plot()

observed(obs_data, x = TIME, y = DV ) |>
  simulated(sim_data, y = DV) |>
  stratify(~ GENDER) |>
  binless(loess.ypc=TRUE) |> 
  predcorrect(pred=PRED) |> #after binless()
  vpcstats() |>
  plot()
```

## tidyvpc 1.3.0

An overview of updates to `plot()` function in `tidyvpc v1.3.0`

----

Set plot output dimensions:

```{r}
knitr::opts_chunk$set(fig.width=12, fig.height=8, dpi = 300) 
```

### One sided stratify() formula uses facet_wrap() {.unlisted .unnumbered}

```{r echo=TRUE, warning=FALSE}
library(tidyvpc)
library(magrittr)
obs_data <- obs_data[MDV == 0]
sim_data <- sim_data[MDV == 0]

vpc <- observed(obs_data, x=TIME, y=DV) %>%
  simulated(sim_data, y=DV) %>%
  stratify(~ GENDER) %>%
  binless() %>%
  vpcstats()

plot(vpc)

```


```{r echo=TRUE, warning=FALSE}
vpc <- observed(obs_data, x=TIME, y=DV) %>%
  simulated(sim_data, y=DV) %>%
  stratify(~ GENDER + STUDY) %>%
  binning(bin = "jenks", nbins = 8) %>%
  vpcstats()

plot(vpc)

```


### Two-sided formula uses facet_grid() {.unlisted .unnumbered}


```{r echo=TRUE, warning=FALSE}
vpc <- observed(obs_data, x=TIME, y=DV) %>%
  simulated(sim_data, y=DV) %>%
  stratify(GENDER ~ STUDY) %>%
  binning(bin = "kmeans", nbins = 6) %>%
  vpcstats()

plot(vpc)

```


### Using facet = TRUE argument {.unlisted .unnumbered}

We can use facet = TRUE argument to facet continuous VPC by quantile or facet categorical VPC by predicted probability.

```{r echo=TRUE, warning=FALSE}
vpc <- observed(obs_data, x=TIME, y=DV) %>%
  simulated(sim_data, y=DV) %>%
  binless() %>%
  vpcstats()

plot(vpc, facet = TRUE, point.alpha = 0.1, point.size = 1, ribbon.alpha = 0.2)

```


```{r echo=TRUE, warning=FALSE}
vpc <- observed(obs_cat_data, x = agemonths, yobs = zlencat) %>%
  simulated(sim_cat_data, ysim = DV) %>%
  binless() %>% 
  vpcstats(vpc.type = "categorical")

plot(vpc, facet = TRUE, legend.position = "bottom")

```


### Changing point size, point alpha, point shape, point stroke, and ribbon alpha {.unlisted .unnumbered}

Setup categorical VPC.

```{r}
vpc <- observed(obs_cat_data, x = agemonths, yobs = zlencat) %>%
  simulated(sim_cat_data, ysim = DV) %>%
  binning(bin = round(agemonths, 0)) %>%
  vpcstats(vpc.type = "categorical")
```

Adjust point size.
```{r}
plot(vpc, point.size = 4)
```


Setup continuous VPC.

```{r echo=TRUE, warning=FALSE}
vpc <- observed(obs_data, x=TIME, y=DV) %>%
  simulated(sim_data, y=DV) %>%
  binless() %>%
  vpcstats()
```

```{r}
plot(vpc, point.size = 1.5, point.stroke = 2.5, point.alpha = 0.1, ribbon.alpha = 0.05)
```

```{r}
plot(vpc, point.size = 5, point.stroke = 0.3, point.shape = "triangle")
```

```{r}
plot(vpc, point.size = 7, point.shape = "square-fill", point.alpha = 0.1, ribbon.alpha = 0.5)

```