---
title: "tidyvpc with RsNLME"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tidyvpc with RsNLME}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, echo=FALSE}
data.table::setDTthreads(2)
```

# Introduction

`tidyvpc` and the `Certara.RsNLME` package integrate well together for a robust pharmacometric workflow. Build and estimate your model using [RsNLME](https://certara.github.io/R-RsNLME/index.html) and easily pass the observed and simulated output `data.frame` to `tidyvpc` for creating visual predictive checks (VPCs). The below steps provide an example workflow when using the `tidyvpc` and `Certara.RsNLME` package together.

# Setup

## R setup

First, you must install the `Certara.RsNLME` package. Complete installation instructions can be found [here](https://certara.github.io/R-RsNLME/articles/installation.html).

```{r setup}
suppressPackageStartupMessages(library(tidyvpc, quietly = TRUE))
library(Certara.RsNLME)
library(ggplot2)
library(magrittr)
```

## Data Exploration 

Let's take a quick look at the time-concentration profile of our input dataset. We will be using the `pkData` `data.frame` from the `Certara.RsNLME` package.

```{r plot-data}
conc_data <- Certara.RsNLME::pkData
conc_data$Subject <- as.factor(conc_data$Subject)
ggplot(conc_data, aes(x = Act_Time, y = Conc, group = Subject, color = Subject)) +
  scale_y_log10() +
  geom_line() +
  geom_point() +
  ylab("Drug Concentration \n at the central compartment")
```

## Model building

The above plot suggests that a two-compartment model with IV bolus is a good starting point. Next, we will define the model using the `numCompartments` argument inside the function `pkmodel()` from the `Certara.RsNLME` package, and provide additional 'column mapping' arguments, which correspond to required model variables mapped to column names in the `conc_data` `data.frame`. 

We will then pipe in additional functions to remove the random effect from `V2`, update initial estimates for fixed and random effects, then change our error model.

```{r model-build}
model <- pkmodel(
  numCompartments = 2,
  data = conc_data,
  ID = "Subject",
  Time = "Act_Time",
  A1 = "Amount",
  CObs = "Conc",
  modelName = "Two-Cmpt") %>%
  structuralParameter(paramName = "V2", hasRandomEffect = FALSE) %>%
  fixedEffect(effect = c("tvV", "tvCl", "tvV2", "tvCl2"),
              value = c(15, 5, 40, 15)) %>%
  randomEffect(effect = c("nV", "nCl", "nCl2"),
               value = rep(0.1, 3)) %>%
  residualError(predName = "C", SD = 0.2)

print(model)
```

## Model fitting

Next, we will fit the model using the function `fitmodel()`.

```{r fit-model, results='hide', warning=FALSE, message=FALSE}
fit_est <- fitmodel(model)
```

## VPC preparation

Now that we have fit the model, we can create a new model with updated parameter estimates and perform our VPC simulation run.

```{r copymodel}
vpc_model <- copyModel(model, acceptAllEffects = TRUE, modelName = "Two-Cmpt-VPC")
print(vpc_model)
```

Next, we will use the `vpcmodel()` function to perform simulation and generate the required observed and simulated input `data.frame` for `tidyvpc`.

```{r vpcsim, results = 'hide', warning = FALSE, message = FALSE}
fit_vpc_sim <- vpcmodel(vpc_model)
```

The resulting observed and simulated data can be found in the returned values from the `vpcmodel()` function e.g.,

```{r get-data}
obs_data <- fit_vpc_sim$predcheck0
sim_data <- fit_vpc_sim$predout
```

## Generate VPC plots

The `x` and `y` arguments to `observed()` are the columns from `fit_vpc_sim$predcheck0`   The `x` and `y` arguments to `simulated()` from `fit_vpc_sim$predout`. By default, `x` column is named `IVAR` and `y` column is named `DV` in the output tables generated from `vpcmodel()`.

```{r vpc-std-setup, warning=FALSE, fig.width = 9, fig.height = 6, out.width=640}
# Create a binless VPC plot
binless_vpc <- observed(obs_data, x = IVAR, yobs = DV) %>%
  simulated(sim_data, ysim = DV) %>%
  binless() %>%
  vpcstats()
plot_binless_vpc <- plot(binless_vpc, legend.position = "none") +
  ggplot2::ggtitle("Binless")

# Create a binning VPC plot with binning method set to be "jenks"
binned_vpc <- observed(obs_data, x = IVAR, yobs = DV) %>%
  simulated(sim_data, ysim = DV) %>%
  binning(bin = "jenks") %>%
  vpcstats()
plot_binned_vpc <- plot(binned_vpc) +
  ggplot2::ggtitle("Binning")

## Put these two plots side-by-side
egg::ggarrange(plot_binned_vpc,
               plot_binless_vpc,
               nrow = 2, ncol = 1,
               top = "VPC Comparison"
)
```
